---
title: "Practical Machine Learning ( Course Project)"
author: "Singappuli Perera"
date: "Friday, March 18, 2016"
output: html_document
---


#Overview
Human Activity Recognition (HAR) is a key research area that is gaining increasing attention, especially for the development of context-aware systems. In this analysis six participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in five different fashions: exactly according to the specification (Class A), throwing the elbows to the front (Class B), lifting the dumbbell only halfway (Class C), lowering the dumbbell only halfway (Class D) and throwing the hips to the front (Class E).

validate the data), to derived the prediction model by using the training data, to validate the model where an expected out-of-sample error rate of less than 0.6%, or 99.4% accuracy, would be acceptable before it is used to perform the prediction on the 20 test cases - that must have 100% accuracy (to obtain 20 points awarded).

#Creating a prediction model

##Loading data

First I will load all of the libraries need for the analyses.
```{r}
setwd("C:/Users/Yamuni/Desktop/Data science specialisation/practical machine learning")

library(lattice)
library(ggplot2)
library(caret)
library(randomForest)
library(e1071)

#download files from the urls provided
#train_url <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
#download.file(url=train_url, destfile="training.csv")

#test_url <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
#download.file(url=test_url, destfile="testing.csv")

#read in training and testing data
train <- read.csv("training.csv", na.strings=c("NA","#DIV/0!",""))
test <- read.csv("testing.csv", na.strings=c("NA","#DIV/0!",""))

#names(train)
#str(train)
#summary(train)
summary(train$classe)#this is the outcome we want to predict

```


#Split training/testing data

Before we do anything, we will set aside a subset of our training data for cross validation (40%). 

#we want to predict the 'classe' variable using any other variable to predict with
```{r}
inTrain <- createDataPartition(y=train$classe, p=0.6, list=FALSE)
myTrain <- train[inTrain, ]
myTest <- train[-inTrain, ]
dim(myTrain)
dim(myTest)

#some exploratory plots
#featurePlot(x=train[, 150:159], y = train$classe, plot = 'pairs')
```

#Feature selection

Now we can tranform the data to only include the variables we will need to build our model. We will remove variables with near zero variance, variables with mostly missing data, and variables that are obviously not useful as predictors.

#first we will remove variables with mostly NAs (use threshold of >75%)

```{r}
mytrain_SUB <- myTrain
for (i in 1:length(myTrain)) {
  if (sum(is.na(myTrain[ , i])) / nrow(myTrain) >= .75) {
    for (j in 1:length(mytrain_SUB)) {
      if (length(grep(names(myTrain[i]), names(mytrain_SUB)[j]))==1) {
        mytrain_SUB <- mytrain_SUB[ , -j]
      }
    }
  }
}

dim(mytrain_SUB)

#names(mytrain_SUB)

#remove columns that are obviously not predictors
mytrain_SUB2 <- mytrain_SUB[,8:length(mytrain_SUB)]

#remove variables with near zero variance
NZV <- nearZeroVar(mytrain_SUB2, saveMetrics = TRUE)
NZV #all false, none to remove

keep <- names(mytrain_SUB2)

```


I decided to use the random forest model to build my machine learning algorithm as it is appropriate for a classification problem as we have and based on information provided in class lectures this model tends to be more accurate than some other classification models.

Below I fit my model on my training data and then use my model to predict classe on my subset of data used for cross validation.

```{r}
#fit model- RANDOM FOREST
set.seed(223)

modFit <- randomForest(classe~., data = mytrain_SUB2)
print(modFit)

#cross validation on my testing data
#out of sample error
predict1 <- predict(modFit, myTest, type = "class")
confusionMatrix(myTest$classe, predict1)
#modFit <- train(classe ~., method = "rf", trControl=trainControl(method = "cv", number = 4), data = mytrain_SUB)
```


#Error

As we can see from the model summaries above, when we run the model on our test data for cross validation we get an accuracy of 99.4% that we can estimate to be our out of sample error. When the model is fitted to the training data used to build the model it shows 100% accuracy, which we can assume as our in sample error.
Apply to final test set

Finally, we apply our model to the final test data. Upon submission all predictions were correct!
```{r}
predict_FINAL <- predict(modFit, test, type = "class")
print(predict_FINAL)

pml_write_files = function(x) {
  n = length(x)
  for (i in 1:n) {
    filename = paste0("problem_id_", i, ".txt")
    write.table(x[i], file=filename, quote=FALSE,row.names=FALSE, col.names=FALSE)
  }
}

pml_write_files(predict_FINAL)
```